Supplementary Report for the article *Functional diversity can facilitate the collapse of an undesirable ecosystem state*

### 2022-11-24 reports/supplement/supplement_0.9.1.Rmd
changes to supplement

Growth of cyanobacteria depended on one substrate (phosphorus), whereas growth of sulfate-reducing bacteria (SB) and phototrophic sulfur bacteria (PB) depended on two substrates (phosphorus and oxidised sulfur for SB, and phosphorus and sulfide for PB).

Inhibition of growth by a substrate was modelled with the Haldane equation:

@@

(ref:temp-meth2) Stable states found with the **temporal method**. All other aspects are the same as in Figure \@ref(fig:repl-meth2).

@@

(ref:div-repl-meth1) **Replication method results.** **Left column of panels**: the effect of trait diversity (y-axis) on the positions of the tipping points (anoxic to oxic in blue, oxic to anoxic in red), which determine the extent of the region of bistability (green shaded region) in the oxygen diffusivity gradient. The left-most region is anoxic only, the middle is oxic or anoxic depending on historical conditions, and the right-most is oxic only. The thick grey vertical lines at -8 and 0 on the x-axis show the extent of the oxygen diffusivity gradient investigated in the simulations. **Right column of panels**: The effect size of trait variation on resilience, with positive values on the y-axis indicating that diversity has increased resilience.

@@

As mentioned in the main report, to avoid very low strain densities, we added 1 unit of density to every biological state variable every 1,000 hours. We here show that a slightly different method produced the same results. Specifically, we tested the effect of making an abundance floor of 1 unit that was implemented every 1,000 hours. I.e. every 1,000 hours, any abundances that were lower than 1 unit were set to 1 unit. As shown in Figure \@ref(fig:div-temp-eventdef1), the results with this floor abundance are similar as when 1 was added every 1,000 hours. Note that a coarser gradient of trait variation was used, hence the saw-tooth patterns. A further difference is the large decrease in resilience of the anoxic to oxic transition above moderate levels of diversity in the SB and CB-SB treatments. This pattern can also occur with addition of 1 abundance unit (e.g., see Figure \@ref(fig:div-temp-meth3)), and is discussed in the section that concerns this below (section [SB diversity effects]).

@@

## Example of transient dynamics at 1e6 hours

The results below show the stable states as a function of oxygen diffusivity for a system with an intermediate level of diversity in all functional groups, with either 1'000'000 hours at each oxygen diffusivity level (Figure \@ref(fig:ss-temp-meth3)), or 2'000'000 hours (Figure \@ref(fig:ss-temp-meth4)). This shows that 1'000'000 hours duration of the simulations are sufficient to reach stable state, except in regions of strain replacement, including at some state transitions. E.g. in the region of strain replacement in the sulfate-reducing bacteria recovery trajectory there is difference in strain abundances at 1'000'000 and 2'000'000 hours. However, the position of the tipping points (which were calculated from the total SB abundance) and the substrate concentrations did not differ between 1'000'000 and 2'000'000 hours.

@@

In the main text, we present results of simulations with nine strains per functional group. Here we also explore the effect of number of strains (2, 3, 6, 9) on the position of the two tipping points (Figure \@ref(fig:div-stab-multstrain1)). At low to moderate amounts of trait variation, the number of strains did not affect the position of the tipping points. However, at high amounts of trait variation, resilience was lower when the functional groups had only two or three strains rather than six or nine. This effect was particularly pronounced when there was high variation in only SB or in SB-CB: here, simulations with two or three strains led to dramatically lower resilience of the anoxic state than simulations with six or nine strains. We believe that this effect is due to the greater distance in trait space between strains in simulations with two or three strains than in simulations with more strains (also see Section 5).

@@

Increased CB diversity causes a slight decrease in the resilience of the anoxic state. The following two figures show the state responses to oxygen diffusivity with low Figure \@ref(fig:ss-CB-only1) and high Figure \@ref(fig:ss-CB-only2) levels of CB diversity. The difference in the position of the anoxic to oxic tipping point is small. It is, in fact the smallest it could be, as it is the smallest step size in oxygen diffusivity used in the simulations.

@@

We showed above that wait times of 1e5 (Figure \@ref(fig:div-temp-meth2)) and 1e4 (Figure \@ref(fig:div-temp-meth4)) produce qualitatively and often quantitatively very similar results to those for wait times of 1e6 (Figure \@ref(fig:div-temp-meth1)). There is, however, a notable difference when there is variation in only SB, or in CB-SB with wait times of 1e4 (Figure \@ref(fig:div-temp-meth4)). Here, moderate increases in diversity have the same effect as with wait times of 1e6. However, larger increases in diversity suddenly cause a large reduction in resilience of the anoxic-oxic transition. Figure \@ref(fig:ss-temp-meth1) gives an example of the stable states when diversity is moderate -- increases in oxygen diffusivity cause strain replacement in the SB to the more tolerant strains, which then delay the shift to the oxic state.  Suprisingly, when there is just a little more diversity in the SB, then the anoxic to oxic transition happens earlier, i.e. there is a large decrease in resilience (Figure \@ref(fig:ss-temp-meth2)).

@@

(ref:pbdiv-only-1) Right-hand panels show the transition from anoxic to oxic state when oxygen diffusivity decreases. The results here are for a **low amount of variation in only the PB functional group**. The transition from anoxic to oxic occurs at approximately $log_{10}$(oxygen diffusivity) = `r round(aO_flip_sim1, 2)`. All other aspects are the same as in Figure \@ref(fig:repl-meth2).

@@

(ref:pbdiv-only-2) Right-hand panels show the transition from anoxic to oxic state when oxygen diffusivity decreases. The results here are for a **moderate amount of variation in only the PB functional group**. The transition from anoxic to oxic occurs at approximately $log_{10}$(oxygen diffusivity) = `r round(aO_flip_sim2, 2)`. All other aspects are the same as in Figure \@ref(fig:repl-meth2).

@@

When there was medium to high variation in only the phototrophic sulfur bacteria (PB), the strain dynamics deviated from the standard pattern: at low oxygen diffusivity, all three functional groups coexisted. Specifically, on the trajectory of decreasing oxygen diffusivity, cyanobacteria reappeared at low oxygen diffusivity after their collapse (Figure \@ref(fig:PB-stabstat1)) or never fully collapsed (high variation in PB, Figure \@ref(fig:PB-stabstat2)). On the trajectory of increasing oxygen diffusivity, cyanobacteria coexisted with the two groups of sulfur bacteria at low oxygen diffusivity, collapsed at medium levels of oxygen diffusivity, and then increased to sole dominance at the tipping point to the oxic state (Figure \@ref(fig:PB-stabstat1)).

>>>>

Inhibition of growth by a substrate was modeled with the Haldane equation:

@@

(ref:temp-meth2) Stable states found with the **temporal method**. The left column of panels refers to the stable state when oxygen diffusivity is at first high (favouring the oxic state). The right column refers to when oxygen diffusivity is at first low (favouring the anoxic state). Points (which sometimes are so close as to appear as thick lines) show the stable state at the end of each period of constant oxygen diffusivity. Thin lines are for visualisation only, and join the points. Grey arrows show the oxygen diffusivity at which the system flips, and the direction of the flip. The length of the grey arrows is without meaning.

@@

(ref:div-repl-meth1) **Replication method results.** **Left column of panels**: the effect of trait diversity (y-axis) on the positions of the tipping points (anoxic to oxic in blue, oxic to anoxic in red), which determine the extent of the region of bistability (green shaded region) in the oxygen diffusivity gradient. The left-most region is anoxic only, the middle is oxic or anoxic depending on historical conditions, and the right-most is oxic only. The thick grey vertical lines at -8 and 0 on the x-axis show the extent of the oxygen diffusivity gradient investigated in the simulations. Data is missing in panels e, and i, when patterns of coexistence were atypical of the classical pattern shown in Figure 1a of the main text (e.g. when cyanobacteria coexisted with sulfur bacteria at low oxygen diffusivities). **Right column of panels**: The effect size of trait variation on resilience, with positive values on the y-axis indicating that diversity has increased resilience.

@@

As mentioned in the main report, to avoid very low strain densities, we added 1 unit of density to every biological state variable every 1,000 hours. We here show that a slightly different method produced the same results. Specifically, we tested the effect of making an abundance floor of 1 unit that was implemented every 1,000 hours. I.e. every 1,000 hours, any abundances that were lower than 1 unit were set to 1 unit. As shown in Figure \@ref(fig:div-temp-eventdef1), the results with this floor abundance are similar as when 1 was added every 1,000 hours.

@@

## 1e6 vs 2e6 hours wait time

The results below show the stable states as a function of oxygen diffusivity for a system with an intermediate level of diversity in all functional groups, with either 1'000'000 hours at each oxygen diffusivity level (Figure \@ref(fig:ss-temp-meth3)), or 2'000'000 hours (Figure \@ref(fig:ss-temp-meth4)). This shows that 1'000'000 hours duration of the simulations are sufficient to reach stable state, except in regions of strain replacement, including at some state transitions. E.g. in the region of strain replacement in the sulfate-reducing bacteria there is difference in strain abundances at 1'000'000 and 2'000'000 hours. However, the position of the tipping points (which were calculated from the total SB abundance) and the substrate concentrations did not differ between 1'000'000 and 2'000'000 hours.

@@

In the main text, we present results of simulations with nine strains per functional group. Here we also explore the effect of number of strains (2, 3, 6, 9) on the position of the two tipping points (Figure \@ref(fig:div-stab-multstrain1)). At low to moderate amounts of trait variation, the number of strains did not affect the position of the tipping points. However, at high amounts of trait variation, resilience was lower when the functional groups had only two or three strains rather than six or nine. This effect was particularly pronounced when there was high variation in only SB or in SB-CB: here, simulations with two or three strains led to dramatically lower resilience of the anoxic state than simulations with six or nine strains. We believe that this effect is due to the greater distance in trait space between strains in simulations with two or three strains than in simulations with more strains (also see Section 9).

@@

Increased CB diversity causes a slight decrease in the resilience of the anoxic state. The following two figures show the state responses to oxygen diffusivity with low (Figure \@ref(fig:ss-CB-only1)) and high (Figure \@ref(fig:ss-CB-only2)) levels of CB diversity. The difference in the position of the anoxic to oxic tipping point is small. It is, in fact the smallest it could be, as it is the smallest step size in oxygen diffusivity used in the simulations.

@@

We showed above that wait times of 1e5 (Figure \@ref(fig:div-temp-meth2)) and 1e4 (Figure \@ref(fig:div-temp-meth3)) produce qualitatively and often quantitatively very similar results to those for wait times of 1e6 (Figure \@ref(fig:div-temp-meth1)). There is, however, a notable difference when there is variation in only SB, or in CB-SB with wait times of 1e4 (Figure \@ref(fig:div-temp-meth3)). Here, moderate increases in diversity have the same effect as with wait times of 1e6. However, larger increases in diversity suddenly cause a large reduction in resilience of the anoxic-oxic transition. Figure \@ref(fig:ss-temp-meth1) gives an example of the stable states when diversity is moderate -- increases in oxygen diffusivity cause strain replacement in the SB to the more tolerant strains, which then delay the shift to the oxic state.  Suprisingly, when there is just a little more diversity in the SB, then the anoxic to oxic transition happens earlier, i.e. there is a large decrease in resilience (Figure \@ref(fig:ss-temp-meth2)).

@@

(ref:pbdiv-only-1) Right-hand panels show the transition from anoxic to oxic state when oxygen diffusivity decreases. The results here are for a **low amount of variation in only the PB functional group**. The transition from anoxic to oxic occurs at approximately $log_{10}$(oxygen diffusivity) = `r round(aO_flip_sim1, 2)`. All other aspects are the same as in Figure \@ref(fig:ss-temp-meth3).

@@

(ref:pbdiv-only-2) Right-hand panels show the transition from anoxic to oxic state when oxygen diffusivity decreases. The results here are for a **moderate amount of variation in only the PB functional group**. The transition from anoxic to oxic occurs at approximately $log_{10}$(oxygen diffusivity) = `r round(aO_flip_sim2, 2)`. All other aspects are the same as in Figure \@ref(fig:ss-temp-meth3).

@@

When there was medium to high variation in only the phototrophic sulfur bacteria (PB), the strain dynamics deviated from the standard pattern: at low oxygen diffusivity, all three functional groups coexisted. Specifically, on the trajectory of decreasing oxygen diffusivity, cyanobacteria reappeared at low oxygen diffusivity after their collapse (medium variation in PB, Figure \@ref(fig:PB-stabstat1)) or never fully collapsed (high variation in PB, Figure \@ref(fig:PB-stabstat2)). On the trajectory of increasing oxygen diffusivity, cyanobacteria coexisted with the two groups of sulfur bacteria at low oxygen diffusivity, collapsed at medium levels of oxygen diffusivity, and then increased to sole dominance at the tipping point to the oxic state (Figure \@ref(fig:PB-stabstat1), Figure \@ref(fig:PB-stabstat2)).
